---
title: "Using_miscPredict_for_evaluating_properties_of_protein_signatures"
author: "Dirk F Moore"
date: "`r Sys.Date()`"
output:
#  word_document:
#    toc: true
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
vignette: |
  %\VignetteIndexEntry{Using_miscPredict_for_evaluating_properties_of_protein_signatures}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
    ),
  options(width = 150),
  eval = TRUE
)
```

\newpage

# 1. Introduction

This vignette provides a guide to using the 'miscPredict' package to 
build a prediction model using a proposed protein signature and evaluate the
predictive accuracy of that model. This vignette requires
a current version of the R programming system, and it is recommended
that you use the RStudio interface to R. Both are available for free at,
respectively,

'https://cran.r-project.org/bin/windows/base/' and
'https://posit.co/download/rstudio-desktop/' (windows version)

An R version for Apple macOS is also available.

As described in the primary reference, Guzman Rivera et al. (2024), 
we have data on the 
relative abundance of 1391 proteins in serum samples for 19 MIS-C 
patients and 40 patients divided among Kawasaki (13patients), 
pneumonia (17 patients), and 
asymptomatic (10 patients). Immunoglobulins were removed from the protein list. 
The protein abundances were measured using the DIA 
(data independent acquisition) mass spectrometry procedure. Our goal is to 
identify protein signatures that distinguish MIS-C patients from the other
disease types and calculate measures of predictive accuracy. 
As explained in Guzman Rivera et al., we propose two three-protein 
signatures and evaluate their properties. We also use signatures
from two previously published papers and evaluate their properties using
our data. In addition, the package described here allows users to
input their own protein signatures to see how well they work using our data.

The 'miscPredict' package takes a proposed protein signature, produces a
predictive model using the support vector machine method, and evaluates 
the resulting model's ability to distinguish MIS-C from the other 
disease types. The evaluation
includes overall error, sensitivity, specificity, and area under the ROC curve
(AUC).

Disclaimer: This package is for research use only.

# 2. Preparing the R system to run the 'miscPredict' package.
The 'miscPredict' package requires that you first download and
install several supporting
packages if you haven't previously installed them. Do this using the
'install' function as follows:

```
install.packages(c("e1071", "pROC", "crossval"))
```

Furthermore, to use the 'miscPredict' package, you must first download 
and install the
'devtools' package and then, from github, the 'miscPredict' package:

```
install.packages("devtools")
devtools::install_github("mooredf22/miscPredict")
```

Next, attach the 'miscPredict' package and the associated data of 
log2-transformed protein abundance levels, and show the first five
rows and selected columns (1, 2, 20, 21, 33, 34, 50 and 51) of the data frame. 
The column names consist of sample type
(M for MIS-C, K for Kawasaki, P for pneumonia, and A for asymptomatic)
followed by the column number (1 through 59). 
The row names are a combination of the gene name and protein identifier.
The values shown
are log2 transformed abundance levels.

```{r, echo=TRUE}
library(miscPredict)
data(logMISCdata)
round(logMISCdata[1:5,c(1:2, 20:21,33:34,50:51 )], digits=2)
```

Finally, we extract the list
of gene names corresponding to the proteins, and list the first five.

```{r, echo=TRUE}
geneList <- matrix(unlist(strsplit(x=rownames(logMISCdata), split ="_")),
                   ncol=2, byrow=TRUE)[,1]
head(geneList)
```

# Obtaining predictions for protein signatures

The first step is to define protein signatures by creating vectors of
the corresponding gene names. Here are the two signatures discussed
in Guzman Rivera et al. (2024) and also one signature from Nygaard (2024)
and one from Yeoh (2024).

```{r, echo=TRUE}
geneSet1 <- c("ORM1", "SERPINA3", "AZGP1")  # Jeisac/Gennaro
geneSet2 <- c("VWF", "FCGBP", "SERPINA3")  # Jeisac/Gennaro

geneSetNygaard <- c("FCGR3A", "LCP1", "SERPINA3", "BCHE") #Nygaard (2024)
geneSetYeoh <- c("CD163", "PCSK9")  #Yeoh (2024)

```

Next, we define a binary variable 'grpInd' which identifies the samples
corresponding to MIS-C and the other samples (Kawasaki, pneumonia, and
asymptomatic). Here is the distribution of the four sample groups:

```{r, echo=TRUE}
grpIndUse <- matrix(unlist(strsplit(x=names(logMISCdata), split="[.]")),
                    ncol=2, byrow=TRUE)[,1]
table(grpIndUse)
```

Since we here focus on comparing MIS-C to the three other groups,
we define an indicator variable which is '1' for MIS-C and '0' for
the other three comparison groups. We see that there are 19 samples 
in the MIS-C group
and 40 in the comparison group:

```{r, echo=TRUE}
grpInd <- as.numeric(grpIndUse == "M")
# 1 indicates MIS-C, 0 indicates Kawasaki, pneumonia, or mild/asymptomatic
table(grpInd)
```

Finally, the 'predictSignature' function uses the signature to
fit a support vector machine 
prediction model and evaluate its predictive accuracy.
We illustrate using geneSet2. A key measure of accuracy is
area under the curve (AUC), obtained as follows:

```{r, echo=TRUE}
set.seed(424020)
pred.Set2 <- predictSignature(genesUse=geneSet2, geneNames=geneList,
                           grpInd=grpInd, log2out=logMISCdata, cv=TRUE)
```

The output data structure 'pred.Set 2' contains a number of components relevant
to the predictive model. The results for area under the curve are 
in the 'out.auc' component:
```{r, echo=TRUE}
pred.Set2$out.auc
```

The model's area under the curve (AUC) is in component 'aucOut' and is 
95.6 percent, and a 95% confidence interval is 89.6% - 100.0%.
The cross-validated estimate of the AUC, which corrects for overfitting, 
is 93.4 percent.

Additional details, including the sensitivity and specificity
of the predictive model and also a table of predicted vs original
numbers of MIS-C and non-MIS-C subjects, may be seen as follows:

```{r, echo=TRUE}
pred.Set2$out.svm
```
To obtain a basic ROC curve, apply the 'plot' function 
to the 'rocAll' svm component:

```{r, echo=TRUE}
plot(pred.Set2$out.svm$rocAll)
```

An ROC curve with a 95 percent confidence interval may be obtained 
by first using the 'ci.se' function in the pROC package to use a bootstrap
calculation method. The 'specificities=seq(0, 1, 0.01)' command specifies a 
list of specificities at which to calculate the confidence limits. Finally,
the 'plot' function creates the plot. The bootstrap method uses randomly
generated bootstrap samples. For reproducibility, one may set the 
random number seed to a specific number.


```{r, echo=TRUE}
set.seed(109540)
sens.ci <- ci.se(pred.Set2$out.svm$rocAll,  
                 boot.n=100, boot.stratified=TRUE,
                 specificities=seq(0, 1, 0.01))
plot(pred.Set2$out.svm$rocAll)
plot(sens.ci, type="shape", col="lightblue")

```


Finally, the results of cross-validation, which corrects for overfitting,
may be obtained thus:

```{r, echo=TRUE}
pred.Set2$out.cv
```


The properties of Nygaard's signature may be obtained by substituting 
in 'geneSetNygaard'

```{r, echo=TRUE}
set.seed(340863)
pred.Set.Nygaard <- predictSignature(genesUse=geneSetNygaard, 
                        geneNames=geneList, grpInd=grpInd, 
                        log2out=logMISCdata, cv=TRUE)
pred.Set.Nygaard$out.auc
```

Additional properties may be obtained by entering 'pred.Set.Nygaard$out.svm' 
and 'pred.Set.Nygaard$out.cv' as described above. 



# References

Nygaard U, Nielsen AB, Dungu KHS, et al. Proteomic profiling reveals 
diagnostic signatures and pathogenic insights in multisystem 
inflammatory syndrome in children. Commun Biol. 2024;7(1):688. 
doi:10.1038/s42003-024-06370-8

Yeoh S, Estrada-Rivadeneyra D, Jackson H, et al. Plasma Protein 
Biomarkers Distinguish Multisystem Inflammatory Syndrome in Children 
From Other Pediatric Infectious and Inflammatory Diseases. 
Pediatr Infect Dis J. 2024;43(5):444-453. doi:10.1097/INF.0000000000004267

# Reproducibility

Following is the output of the utility `sessionInfo`. 
This output contains details of the packages and version numbers 
used to generate these tutorials.


```{r, echo=TRUE, eval=TRUE}
print(utils::sessionInfo(), width=80)
```
