---
title: "Using miscPredict for evaluating properties of protein signatures"
author: "Dirk F Moore"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
vignette: |
  %\VignetteIndexEntry{Using miscPredict for evaluating properties of protein signatures}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
    ),
  options(width = 150),
  eval = TRUE
)
```

\newpage

# 1.Introduction

As described in the primary reference, Guzman Rivera et al. (2024), 
we have data on the 
relative abundance of 1391 proteins in serum samples for 19 MIS-C 
patients and 40 patients divided among Kawasaki (13patients), 
pneumonia (17 patients), and 
asymptomatic (10 patients). Immunoglobulins were removed from the protein list. 
The protein abundances were measured using the DIA 
(data independent acquisition) mass spectrometry procedure. Our goal is to 
identify protein signatures that distinguish MIS-C patients from the other
disease types and calculate measures of predictive accuracy. 
As explained in Guzman Rivera et al., we propose two three-protein 
signatures and evaluate their properties. We also use signatures
from two previously published papers and evaluate their properties using
our data. In addition, the package described here allows users to
input their own protein signatures to see how well they work using our data.

The 'miscPredict' package takes a proposed protein signature, produces a
predictive model using the support vector machine method, and evaluates 
the resulting model's ability to distinguish MIS-C from the other 
disease types. The evaluation
includes overall error, sensitivity, specificity, and area under the ROC curve
(AUC).

# 2. Setting up the data for 'miscPredict'.
To use the 'miscPredict' package, first download and install the
'devtools' package and then, from github, the 'miscPredict' package:

```
install.packages("devtools")
devtools::install_github("mooredf22/miscPredict")
```

Next, attach the 'miscPredict' package and the associated data of 
log2-transformed protein abundance levels, and show a few
selected rows and columns of the data frame. 
The column names consist of sample type
(M for MIS-C, K for Kawasaki, P for pneumonia, and A for asymptomatic)
followed by the column number (1 through 59). The values shown
are log2 transformed abundance levels.

```{r, echo=TRUE}
library(miscPredict)
data(logMISCdata)
round(logMISCdata[1:5,c(1:2, 20:21,33:34,50:51 )], digits=2)
```

Finally, we extract the list
of gene names corresponding to the proteins:

```{r, echo=TRUE}
geneList <- matrix(unlist(strsplit(x=rownames(logMISCdata), split ="_")),
                   ncol=2, byrow=TRUE)[,1]
```

# Obtaining predictions for protein signatures

The first step is to define protein signatures by creating vectors of
the corresponding gene names. Here are the two signatures discussed
in Guzman Rivera et al. (2024):

```{r, echo=TRUE}
genesSet1 <- c("ORM1", "SERPINA3", "AZGP1")  # Jeisac/Gennaro
genesSet2 <- c("VWF", "FCGBP", "SERPINA3")  # Jeisac/Gennaro

genes4Nygaard2024 <- c("FCGR3A", "LCP1", "SERPINA3", "BCHE") #Nygaard2024
genesYeoh <- c("CD163", "PCSK9")

```

Next, we define a binary variable 'grpInd' which identifies the samples
corresponding to MIS-C and the other samples (Kawasaki, pneumonia, and
asympatomatic). We see that there are 19 samples in the MIS-C group
and 40 in the comparison group.

```{r, echo=TRUE}

grpIndUse <- matrix(unlist(strsplit(x=names(logMISCdata), split="[.]")),
                    ncol=2, byrow=TRUE)[,1]
grpInd <- as.numeric(grpIndUse == "M")
# 1 indicates MIS-C, 0 indicates Kawasaki, pneumonia, or mild/asymptomatic
table(grpInd)
```
Finally, the 'predictSignature' function uses the signaure (here 'geneSet2') to
fit a support vector machine 
prediction model and evaluate its predictive accuracy.

```{r, echo=TRUE}
pred.Set2 <- predictSignature(genesUse=genesSet2, geneNames=geneList,
                           grpInd=grpInd, log2out=logMISCdata, cv=TRUE)
pred.Set2$out.svm
pred.Set2$out.cv
```
The model's area under the curve (AUC) is in component 'aucOut' and is 
95.6 percent, and a 95% confidence interval is 89.6% - 100.0%.
The cross-validated estimate of the AUC, which corrects for overfitting, 
is 93.4 percent.

The properties of Nygaard's signature may be obtained by substituting 
in 'genes4Nygaard2024'

```{r, echo=TRUE}
pred.Set2 <- predictSignature(genesUse=genes4Nygaard2024, geneNames=geneList,
                           grpInd=grpInd, log2out=logMISCdata, cv=TRUE)
pred.Set2$out.svm
pred.Set2$out.cv
```
